Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data

### 2014-01-23 doc/2014bioinformatics_zhang/RVD3_1.tex

\section{Motivation:}
Including prior information in the Bayesian approach is difficult but meaningful.

\section{Results:}
We analyzed the Sensitivity of the Bayesian model to different priors. Jeffreys prior improves robustness and accuracy of Single Nucleotide Variant detection algorithm.

\section{Availability:}

@@

\section{Introduction}
Massively parallel sequencing data were generated by Next-generation sequencing (NGS) technology to benefit clinical diagnostics and sequencing based phylogenetic analyses. One primary application of NGS is variation detection among related populations and separate novel single-nucleotide variants (SNV) candidates.

Currently SNV are deduced on NGS platform by variants calling algorithms. Typically, SAMtools, Genome Analysis Toolkit (GATK) \citep{McKenna:2010bva}, VarScan2 \citep{Koboldt:2012cga}, Strelka \citep{Saunders:2012fh}, and muTect \citep{Cibulskis:2013ta} are all highly used in detecting low-frequency variants. However identifying true SNVs remains challenging because of high false positive rates or high false discovery rate. Another statistical method-RVD \citep{Flaherty:2011ja}, was proposed based on beta-binomial model for ultrasensitive rare single nucleotide variant detection. So based on the RVD hierarchical model we built an improved robustness and accuracy model for low-depth variants detection.

%%%%%%%%%%%%%%%%%%
% RVD3 Model
%%%%%%%%%%%%%%%%%%
\section{RVD3 Model}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
% Priors for precision parameter
%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Priors for precision parameter}
%\subsubsection{Improper Prior}
The prior distribution characterizes the knowledge of the parameters in the model. An improper prior is the prior distribution integrates to infinity, and may cause an improper posterior which results in an invalid inferences \citep{lesaffre2012bayesian}. Furthermore when the Markov chain Monte Carlo method is taken to derive the posterior, it is possibly hard to sniff out the improper posterior. Even though no problems happened in estimation by improper prior, other troubles could be caused in the Bayesian inference and analysis by it \citep{stein1965approximation}. Playing no prior on $ M_{j}$ is exactly an implicit improper prior. Therefore we considered about non-information prior and information prior.

@@

We demonstrated our RVD3 model with Jeffreys prior on yeast data to identify the variants.

\section{Discussion}

Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text. Figure \ref{fig:02} shows that the above method  Text

@@

(Table~\ref{Tab:01}) Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text. Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.  \citealp{Boffelli03} might want to know about  text text text text

@@

\paragraph{Funding\textcolon} Text Text Text Text Text Text  Text Text.

>>>>

\section{Motivation:}
Massively parallel sequencing data has been generated by next-generation sequencing (NGS) technology for single nucleotide variants (SNVs) identification among related populations. To address the detection of SNVs, a variety of methods are being under-represented. However, by the reason of the error rate of the clonal heterogeneity and the limitation of the algorithms, identifying the true variants at minor allele frequencies for the very low sequencing read depth remains challenging.

\section{Results:}
We propose a novel empirical Bayesian model to call variants accurately. And the sensitivity of this hierarchical model is analyzed by the improper prior, information prior and non-information prior on the synthetic sequence data with varying read depth and a range of allele frequencies. Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39). Moreover, the model with Jeffreys prior presents much lower false discovery rate (FDR) to a known 0.1\% minor allele frequency event, compared with using improper prior. For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth. Thus our Bayesian model achieves a enhanced robustness and accuracy when calling variants for the low read depth and minor allele frequencies.
\section{Availability:}

@@

\section{Introduction}
Massively parallel sequencing data has been generated by Next-generation sequencing (NGS) technology to benefit clinical diagnostics and sequencing based phylogenetic analyses.
One primary application of NGS is variation detection among related populations and separate novel single-nucleotide variants (SNVs) candidates.
Somatic SNVs are detected by comparing the tumor and corresponding normal samples.

To address the detection of SNVs at low allele frequencies, a number of algorithms from NGS platform are being under-represented.
Strelka \citep{Saunders:2012fh}, VarScan2 \citep{Koboldt:2012cga}, JointSNVMix \citep{roth2012jointsnvmix} are highly used to differentiate somatic SNVs from germline cells.
Also, SAMtools, Genome Analysis Toolkit (GATK) \citep{McKenna:2010bva}, and muTect \citep{Cibulskis:2013ta} broadly concentrate on detecting low-frequency variants.
However identifying the true SNVs remains challenging because of the high false positive rate or high false negative rate which are mainly caused by the clonal heterogeneity.

Recently empirical Bayesian approaches have been made use to identify SNVs.
A empirical Bayesian framework for somatic mutation detection from cancer genome sequencing data - EBCall, enables accurate mutations calling with low allele frequencies (less than 10\%) in a minor tumour subpopulation \citep{shiraishi2013empirical}.
Another method based on empirical Bayesian hierarchical model - RVD, was proposed for ultrasensitive rare SNV detection using beta-binomial model \citep{Flaherty:2011ja}. RVD method is demonstrated to robustly detect mutations at 0.1\% fractional representation, which means accurately call one mutant per every 1000 wild-type alleles.
With the shortcoming of high read depth estimation in RVD, we originally built an improved robustness and accuracy model - RVD3 for low-depth SNVs detection. Variants calling ability is tested and analyzed both on the synthetic sequence data and the true yeast sequencing data with different read depths.

In this article, RVD3 model - a novel Bayesian structure to accurately identify SNVs with small false discovery rate is first described in detail.
Secondly, Metropolis-within-Gibbs sampling is evolved for inference. And then to detect variants, a Bayesian posterior distribution is taken for hypothesis test.
Furthermore, we analyze the sensitivity of the Bayesian model to different priors - Jeffreys prior and log-normal prior, and apply the different priors on the synthetic sequence data.
Finally, we choose Jeffreys prior for the RVD3 model and demonstrate its performance on the yeast sequence data.
The analysis for the results reveals the enhanced robustness and improved accuracy of our algorithm.
%%%%%%%%%%%%%%%%%%
% RVD3 Model
%%%%%%%%%%%%%%%%%%
\section{RVD3 Model}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
% Priors for precision parameter
%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Priors for precision parameter}
%\subsubsection{Improper Prior}
The prior distribution characterizes the knowledge of the parameters in the model. Including prior information in the Bayesian approach is difficult but meaningful. An improper prior is the prior distribution integrates to infinity, and may cause an improper posterior which results in an invalid inferences \citep{lesaffre2012bayesian}. Furthermore when the Markov chain Monte Carlo method is taken to derive the posterior, it is possibly hard to sniff out the improper posterior. Even though no problems happened in estimation by improper prior, other troubles could be caused in the Bayesian inference and analysis by it \citep{stein1965approximation}. Playing no prior on $ M_{j}$ is exactly an implicit improper prior. Therefore we considered about non-information prior and information prior.

@@

We demonstrated our RVD3 model with Jeffreys prior on yeast data to identify the variants \citep{Flaherty:2011ja}.

\section{Discussion}
analysis of the tradeoff between experimental precision and sequencing depth to achieve ultrasensitive variant detection.

@@

@@

\paragraph{Funding\textcolon} Text Text  Text Text.
